%! TeX program = lualatex
\documentclass[twoside, a4paper, 12pt]{book}

\title{Algebraic geometry}
\author{\color{subtext1}a desperate attempt to avoid failure}
\date{}

\usepackage[pl, dates]{../../../../template}

\begin{document}
\frontmatter 
\maketitle
\thispagestyle{empty}
\setcounter{page}{0}
%
% \tableofcontents

\mainmatter

\pagestyle{fancy}

 
% \include{./rozdzialy/01_schemes/01_schemes.tex} 


% \include{czesci/01_wstep.tex}
%
% \include{czesci/02.tex}

\chapter{Blitzkrieg z GA}

\section{}{Noetherowskość, wymiary}

For any $A\subseteq K[\overline{X}]$ we define
$$V(A)=\{\overline{x}\in K^n\;:\;\forall\;F\in A\;F(\overline{x})=0\}.$$
Let $I,J\trianglelefteq K[\overline{X}]$. Then
\begin{itemize}
  \item $A_0\subseteq A_1\implies V(A_1)\subseteq V(A_0)$
  \item $V(\bigcup A_i)=\bigcap V(A_i)$
  \item $V(I\cap J)=V(IJ)=V(I)\cup V(J)$
  \item $V(I+J)=V(I)\cap V(J)$
\end{itemize}

For any $A\subseteq K[\overline{X}]$ there is a finite $A_0\subseteq A$ such that $V(A)=V(A_0)$ (by the Hilbert's basis theorem).

\begin{definition}{Noetherian ...}{}
  A topological space $X$ is called \buff{Noetherian} if any descending chain of closed subsets of $X$ stabilizes. Meanwhile, a ring is Noetherian if any ascending chain of ideals stabilizes.
\end{definition}

\begin{definition}{irreducible space}{}
  A space $X$ is irreducible if it is not a non-trivial union of its two closed subseteq, i.e. for any $Y_1,Y_2\subseteq X$ closed $X=Y_1\cup Y_2$ then $X=Y_1$ or $Y_2$.
\end{definition}

If $X$ is a Noetherian space then
\begin{itemize}
  \item $X=X_1\cup...\cup X_k$ for some $X_1,...,X_k\subseteq X$ irreducible such that $X_i\not\subseteq X_j$ for $i\neq j$
  \item this sequence is unique up to permutation of indices
\end{itemize}

An affinie algebraic set $V(A)$ is \buff{affine variety} if it is irreducble as a topological space with the Zariski topology.

\begin{definition}{dimension}{}
  $\dim(X)=n$ if there is a strictly decreasing sequence of irreducible closed subsets of $X$ such that 
  $$X_n\subsetneq X_{n-1}\subsetneq...\subsetneq X_0\subseteq X$$
\end{definition}

For $V\subseteq \mathbb{A}^n$ we define the \buff{affine coordinate ring of $V$} as
$$K[v]:=\{f\in \Func(V,K)\;:\;\exists F\in K[\overline{X}]\;:F|_V=f\}$$
$K(V)$ is the field of fractions of $K[V]$, called \buff{the field of rational functions on $V$}.

The \buff{ideal of $V$} is defined as 
$$\ker\left(K[\overline{X}]\ni F\mapsto F|_V\in K[V]\right)$$
which is the same as
$$I(V)=\{F\in K[\overline{X}]\;:\;\forall\;\overline{x}\in V\;F(\overline{x})=0\}.$$
The \buff{Zariski closure} of $V_0$ is the set $V(I(V_0))$.

\begin{theorem}{Hillbert's Nullstellensatz}{}
  \begin{description}
    \item[weak] $I\trianglelefteq K[\overline{X}]\;\land;I\neq K[\overline{X}]\implies V(I)\neq\emptyset$
    \item[strong/regular] $I\trianglelefteq K[\overline{X}]\implies I(V(I))=\sqrt{I}$
  \end{description}
\end{theorem}

If $V\subseteq \mathbb{A}^n$ is Zariski closed, then the following are equivalent
\begin{enumerate}
  \item $V$ is irreducible
  \item $I(V)$ is prime
  \item $\exists P\trianglelefteq K[\overline{X}]$ prime such that $V=V(P)$ !!the field $K$ needs to be algebraically closed!!
  \item $K[V]$ is a domain
\end{enumerate}

\begin{theorem}{}{}
  If $F\in K[\overline{X}]$ is irreducible, then $V(F)$ is an affine variety.
\end{theorem}

Let $V\subseteq\mathbb{A}^n$ be an affine algebraic set. Then there is a bijection between $\begin{matrix}\text{radical}\\\text{prime}\end{matrix}$ ideals of $K[V]$ and the set of Zariski $\begin{matrix}\text{closed}\\\text{irreducible closed}\end{matrix}$ subsets of $V$.

\begin{definition}{Krull dimension}{}
  For a ring $R$ we define its Krull dimension $\dim(R)$ as the supremum $k\in\N$ such that there is a strictly increasing (or decreasing) sequence of prime ideals
  $$P_0\subsetneq P_1\subsetneq ...\subsetneq P_k$$
  of $R$.
\end{definition}
{\large\boldmath$$\color{blue}\dim(V)=\dim(K[V])$$}

\begin{theorem}{}{}
  Let $R$ be a finitely generated $K$-algebra which is a domain. Then
  $$\dim(R)=\text{trdeg}_K(R_0)$$
  the dimension of $R$ is equal to the transcendental degree of the field of fractions of $R$ over $K$.
\end{theorem}

If $R=K[V]$ then $R_0=K(V)$ and the above statement holds.

\section{}{Kategoryje}

Let $V\subseteq \mathbb{A}^n$ and $W\subseteq \mathbb{A}^m$ be affine algebraic sets. Then $\phi:V\to W$ is a \buff{morphism} if there are $f_1,...,f_m\in K[V]$ such that
$$\phi(v)=(f_1(v),...,f_m(v)).$$

For such a morphism we define
$$\phi^*:K[W]\to K[V]$$
$$\phi^*(f)=f\circ \phi.$$
\begin{enumerate}
  \item The mapping $\phi\mapsto \phi^*$ is an isomorphism between the set of morphisms $V\to W$ and the set of morphisms $K[W]$ to $K[V]$.
  \item Any finitely generated $K$-algebra $R$ which is reduced (no nilpotent elements) is isomorphic over $V$ to $K[V]$ for some affine algebraic $V$.
\end{enumerate}

{\color{blue}\large\boldmath
$$V\cong W\iff K[V]\cong_K K[W]$$
}

For $f\in K(V)$, the \buff{domain} of $f$, $\dom f$, is the set of points $v\in V$ such that there are $f_1, f_2\in K[V]$, $f=\frac{f_1}{f_2}$ and $f_2(v)\neq0$.

\begin{definition}{}{}
  Let $f\in K(V)$ and $v\in V$.
  \begin{enumerate}
    \item $f$ is regular at $v$ if $v\in \dom f$
    \item $\Oo_{V,v}:=\{f\in K(V)\;:\;v\in \dom f\}$
    \item $f$ is regular if $f$ is regular at each $v\in V$, i.e. $\dom f=V$.
  \end{enumerate}
\end{definition}

\begin{fact}{}{}
  For any $v\in V$
  $$\Oo_{V,v}=K[V]_{I_V(v)}=\{\frac{a}{b}\;:\;a\in K[V],b\in K[V]-I_V(v)\}$$
\end{fact}

Denote by
$$\mm_{V,v}\trianglelefteq \Oo_{V,v}$$
the maximal ideal of $\Oo_{V,v}$.

$f$ is regular $\iff$ $f\in K[V]$

\begin{lemma}{}{}
  For a morphism $\phi:V\to W$ its dual $\phi^*$ is a monomorphism $\iff$ $\phi$ is \buff{dominant}, i.e. $\phi(V)$ is Zariski dense in $W$ (is an epimorphism in its category).
\end{lemma}

A function $\phi:U\subseteq V\to W$ is a \buff{rational function} between $V$ and $W$ if there are $f_1,..., f_m\in K(V)$ such that 
$$U=\dom f_1\cap...\cap \dom f_m$$
and for all $v\in U$ there is $\phi(v)=(f_1(v),..., f_m(v))$.

$\phi:V\dasharrow W$ denotes a \buff{dominant rational function} from $V$ to $W$.

For any field extension $K\subseteq L$ such that $L$ is finitely generated over $K$ there is an affine variety $V$ such that $L\cong_K K(V)$.

\textbf{The category of affine varieties and dominant rational maps is \emph{entiequivalent} or \emph{dually equivalent} to the category of finitely generated field extensions of $K$}.

\section{}{Smooooth \emph{like the fur of a newborn goat}}

Let $R$ be a ring. The map $\partial R\to R$ is called a \buff{derivation} on $R$ if for all $a,b\in R$ 
$$\partial(a+b)=\partial(a)+\partial(b)$$
$$\partial(ab)=\partial(a)b+a\partial(b).$$

The \buff{Jacobian matrix} of $\overline{F}=(F_1,...,F_m)$, $F_i\in K[\overline{X}]$ is 
$$J_{\overline{F}}:=\begin{pmatrix}\frac{\partial F_1}{\partial X_1} & \dots & \frac{\partial F_1}{\partial X_n}\\\vdots & \ddots & \vdots \\ \frac{\partial F_m}{\partial X_1}&\dots &\frac{\partial F_m}{\partial X_n}\end{pmatrix}$$

\begin{fact}{}{}
  If $(G_1,..., G_k)=I=(F_1,..., F_m)\trianglelefteq K[\overline{X}]$ and $v\in V(I)$, then 
  $$\rank(J_{\overline{G}}(v))=\rank(J_{\overline{F}}(v)).$$
\end{fact}

\begin{definition}{non-singular variety}{}
  Let $V\subseteq\A^n$ and $F_1,..., F_m\in I(V)=(F_1,..., F_m)$. We say that $a\in V$ is a \buff{non-singular} or smooth point of $V$ if 
  $$\rank(J_{\overline{F}}(a))=n-\dim(V).$$
  We say that $V$ is a non-singular variety or a smooth variety if $V$ is irreducible and all points of $V$ are smooth.
\end{definition}

$F\in K[X,Y]$ and $V=V(F)\subseteq \A^2$
\begin{enumerate}
  \item $F\notin K\implies |V|=\infty$
  \item $|V(F, \frac{\partial F}{\partial X}, \frac{\partial F}{\partial Y})|<\infty\implies\sqrt{(F)}=(F)\;\land\;I(V)=(F)$
  \item $V(F, \frac{\partial F}{\partial X}, \frac{\partial F}{\partial Y})=\emptyset \implies V$ is smooth 
\end{enumerate}

\begin{lemma}{}{}
  A point $a\in V$ is smooth $\iff$ $\dim_K(I_V(a)/I_V(a)^2)=\dim(V)$.

  If $V$ is an affine variety and $a\in V$, then we have
  $$I_V(a)/I_V(a)^2\cong_K\mm_{V,a}/\mm_{V,a}^2$$
\end{lemma}

% $(R, \mm)$ is a DVR and $r\in R$. 

A Noetherian local ring $(R, \mm)$ is \buff{regular} if $\dim(R)=\dim_{R/\mm}(\mm/\mm^2)$.

The $K$-vector space $\mm_{V,a}/\mm_{V,a}^2$ is the \buff{cotangent space} of $V$ at $a$. The dual space is the \acc{tangent space}.

\section{}{DVR}

A local ring $(R,\mm)$ is a \buff{discrete valuation ring} if
\begin{enumerate}
  \item $R$ is Noetherian domain
  \item $R$ is not a field
  \item $\mm$ is principal (generated by a single element)
\end{enumerate}

In any Noetherian domain $R$ and any $I\triangleleft R$ we have
$$\bigcup_{n\geq 1}I^n=\{0\}.$$

{\color{blue}\large\bfseries\centering Any DVR is PID (the generator of $\mm$ is the uniformizing parameter)}

Let $R$ be a UFD, $r\in R$ irreducible and $R_0$ be the field of fractions of $R$. Define
$$v_r:R_0^*\to \Z$$
$$v_r(r^n\frac{a}{b})=n,\quad r\not| a,\;r\not| b.$$
We call $v_r$ the \buff{r-addic valuation} on $R_0$.

\begin{fact}{}{}
  $R$, $r$, $R_0$ and $v_r$ as above. Then for all $\alpha, \beta\in R_0^*$
  \begin{enumerate}
    \item $\alpha+\beta\in R_0^*\implies v_r(\alpha+\beta)\geq \min\{v_r(\alpha), v_r(\beta)\}$
    \item $v_r(\alpha\beta)=v_r(\alpha)+v_r(\beta)$
    \item $v_r(R_0^*)=\Z$
  \end{enumerate}
\end{fact}

For any irreducible $r,s\in R$ if $(r)=(s)$ then $v_r=v_s$.

\begin{definition}{discrete valuation}{}
  Let $L$ be a field. Any function $v:L^*\to \Z$ satisfying 1-3 from the fact above is called a \buff{(discrete) valuation} on $L$. For any valuation $v:L^*\to\Z$ 
  \begin{itemize}
    \item $\Oo_v:=\{\alpha\in L^*\;:\; v(\alpha)\geq 0\}\cup \{0\}$ -> \acc{valuation ring} of $v$
    \item $\mm_v:=\{\alpha\in L^*\;:\; v(\alpha)>0\}\cup \{0\}$ -> \acc{valuation ideal} of $v$
  \end{itemize}
\end{definition}

For a valuation $v:L^*\to \Z$, $(\Oo_v,\mm_n)$ is a DVR.

\begin{theorem}{}{}
  Let $C$ be an affine curve and $a\in C$. Then $a$ is smooth $\iff$ $(\Oo_{C,a},\mm_{C,a})$ is a DVR.
\end{theorem}

\begin{definition}{}{}
  Let $C$ be an affine curve and $a\in C$ be a smooth point
  \begin{enumerate}
    \item a uniformizing parameter $f\in\Oo_{C,a}$ is a \buff{local parameter} for $C$ at $a$
    \item the unique valuation on $K(C)$ given by $(\Oo_{C,a},\mm_{C,a})$ is denoted $\ord_a$
    \item for $f\in K(C)-\{0\}$ and $n\in\N_{>0}$
      \begin{itemize}
        \item $\ord_a(f) =n\implies $ $f$ has a zero at $a$ of order $n$
        \item $\ord_a(f)=-n\implies$ $r$ has a pole at $a$ of order $n$
      \end{itemize}
  \end{enumerate}
\end{definition}
$$\ord_a(f)=\dim_K(\Oo_{C,a}/f\Oo_{C,a})$$
$$\dim_K(K[X]/(F))=\sum_{a\in V(F)}\ord_a(F)$$

\begin{definition}{intersection number}{}
  The intersection number of $F$ and $G$ at $a=(x,y)\in \A^2$ is
  $$I(a, F\cap G):=\dim_K(\Oo/(F,G)\Oo),$$
  where 
  $$\Oo:=K[X,Y]_{(X-x, Y-y)}=K[X,Y]_{I(a)}=\Oo_{\A^2,a}$$

  For curves $C_1, C_2$ such that $F=I(C_1)$ and $G=I(C_2)$ we define $I(a, C_1\cap C_2):=I(a, F\cap G)$.
\end{definition}

$$I(a, F\cap G)>0\iff a\in V(F,G)$$
$$|V(F,G)|<\infty\implies I(a, F\cap G)<\infty$$
\begin{center}
  $F$ irreducible and $a\in V(F)$ smooth then 
  $$I(a, F\cap G)=\ord_a(G|_{V(F)}).$$
\end{center}

26/43



\end{document}
